Some lessons after the formalization of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-05-30 Independence_CH/non-absolute.tex
Minor editing

\subsection{Tools for relativization}
\label{sec:tools-relativization}
In order to cope with this, we added the missing step from the
literature consisting of relative versions of the various non-absolute
functions and programmed in ML limited automatic facilities that
define the needed concepts, and  state and
prove the requisite lemmas. For instance, the 
$\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
\session{Isabelle/ZF}, and the commands
\begin{isabelle}
  \isacommand{relativize}\isamarkupfalse%
  \ \isakeyword{functional}\ {\isachardoublequoteopen}cardinal{\isachardoublequoteclose}\ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ \isakeyword{external}\isanewline
  \isacommand{relationalize}\isamarkupfalse%
  \ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\isanewline
  \isacommand{synthesize}\isamarkupfalse%
  \ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}{\kern0pt}definition}\ \isakeyword{assuming}\ {\isachardoublequoteopen}nonempty{\isachardoublequoteclose}%
\end{isabelle}
define the relative cardinal function
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
(denoted  $|\cdot|^M$, as expected),
its relational version $\isatt{is{\uscore}cardinal}$, the
internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
satisfaction by a set is equivalent to the relational version, and
prove the previous statement (analogous to (\ref{eq:sats_big_union_fm})).
The proof that $\isatt{is{\uscore}cardinal}(M,x,z)$  encodes the
statement $|x|^M = z$ must still be done by hand, since the definition
of $\isatt{cardinal{\uscore}rel}$ already involves some tacit
absoluteness results (“\textit{the least $z \in \Ord$ such that $z
  \approx^M x$}” instead
of “\textit{the least $z \in \Ord^M$ such that $z
  \approx^M x$}”, and the like).

>>>>

\subsection{Tools for relativization}
\label{sec:tools-relativization}
In order to cope with this, we added the missing step from the
literature consisting of relative versions of the various non-absolute
functions and programmed in ML limited automatic facilities that
define the needed concepts, and  state and
prove the requisite lemmas. For instance, the 
$\isatt{cardinal}::\tyi \fun \tyi$ function is defined in
\session{Isabelle/ZF}, and the commands
\begin{isabelle}
  \isacommand{relativize}\isamarkupfalse%
  \ \isakeyword{functional}\ {\isachardoublequoteopen}cardinal{\isachardoublequoteclose}\ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ \isakeyword{external}\isanewline
  \isacommand{relationalize}\isamarkupfalse%
  \ {\isachardoublequoteopen}cardinal{\isacharunderscore}{\kern0pt}rel{\isachardoublequoteclose}\ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\isanewline
  \isacommand{synthesize}\isamarkupfalse%
  \ {\isachardoublequoteopen}is{\isacharunderscore}{\kern0pt}cardinal{\isachardoublequoteclose}\ \isakeyword{from{\isacharunderscore}{\kern0pt}definition}\ \isakeyword{assuming}\ {\isachardoublequoteopen}nonempty{\isachardoublequoteclose}%
\end{isabelle}
define the relative cardinal function
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
(denoted  $|\cdot|^M$, as expected),
the relational version of the latter $\isatt{is{\uscore}cardinal}$, the
internalized formula \isatt{is{\uscore}cardinal{\uscore}fm} whose
satisfaction by a set is equivalent to the relational version, and
prove the previous statement (analogous to (\ref{eq:sats_big_union_fm})).
The proof that $\isatt{is{\uscore}cardinal}(M,x,z)$  encodes the
statement $|x|^M = z$ must still be done by hand, since the definition
of $\isatt{cardinal{\uscore}rel}$ already involves some tacit
absoluteness results (“\textit{the least $z \in \Ord$ such that $z
  \approx^M x$}” instead
of “\textit{the least $z \in \Ord^M$ such that $z
  \approx^M x$}”, and the like).

### 2022-05-30 Independence_CH/proof_assists_isabelle.tex
Minor editing

The type of propositions $\prop$ is equipped with a binary operation
${\implies}$ (“meta-implication”) and a universal “meta-quantifier”
$\ALL$, that are used to represent the object
logic rules. As an example, the axiomatization of first-order logic
postulates a type $\tyo$ of booleans, and Modus Ponens
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#IFOL.mp|axiom
is written as
\[
  \ALL P\,Q.\ \ [P\limp Q] \implies ([P] \implies [Q]).
\]
The square brackets (which are omitted in Isabelle theories) represent
an injection from $\tyo$ into $\prop$. % ($[P]$ can be read as “$P$
% holds”)
A consequence of this representation is that every formula of
the object logic appears atomic to $\Meta$.

@@

Paulson  \cite{Paulson1989} carried out a proof that the encoding
$\Meta_{\mathrm{IFOL}}$ of
intuitionistic first-order logic IFOL without equality  in the original $\Meta$ is
conservative (there is a correspondence between provable $\phi$ in
IFOL and provable $[\phi]$ in $\Meta_{\mathrm{IFOL}}$) by putting
$\Meta_{\mathrm{IFOL}}$ proofs in \emph{expanded normal form}
\cite{MR0387024}. Passing to classical logic does not present
difficulties, but the addition of meta-equality must be taken care of.
Even more so, since the treatment of equality differs between the
original and the present incarnation of $\Meta$; details for the
latter are exhaustively expounded in the recent formalization by
Nipkow and Roßkof \cite{10.1007/978-3-030-79876-5_6}.

@@

For the most part, the development of set theory in Isabelle is
carried out using its ZF object logic
\cite{DBLP:journals/jar/PaulsonG96}, which is the third logical layer
of the formalization and the most versatile one, since 
Isabelle's native automation is available at this level. Apart from
the type and sort
declarations detailed above, it features a finite axiomatization,
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.mem|const
with a predicate for membership, constants for the empty set and an
infinite set, and functions $\isatt{Pow}::\tyi\fun\tyi$,
$\union::\tyi\fun\tyi$, and $\isatt{PrimReplace} :: \tyi \fun (\tyi
\fun \tyi \fun\tyo)\fun \tyi$ (for Replacement). The Axiom of
Replacement
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.replacement|axiom
has a free predicate variable $P$: % $P ::\tyi\fun\tyi\fun\tyo$:
\begin{multline*}
  (\forall x \in A .\ \forall y\, z.\ P(x, y) \wedge P(x, z)
  \longrightarrow y=z) \implies \\
  b \in \isatt{PrimReplace}(A, P)
  \longleftrightarrow(\exists x \in A .\ P(x, b)) 
\end{multline*}
The restrictions on sorts described above ensure that it is not
possible that higher-order quantification gets used in $P$. The
statement of $\AC$ also has a higher-order free variable, but we only
use $\AC$ for demonstration purposes.

@@

The development of relativization and absoluteness for classes $C::
\isatt{i} \fun \isatt{o}$ follows the same pattern. Each particular
concept was manually written in a relational form and relativized.
Here, the contrast between the usual way one regards $\ZF$ as
first-order theory in the language $\{\in \}$ and the mathematical
practice of freely using defined concepts comes to the
forefront. Assistants have refined mechanisms to cope with defined
concepts (which also make their foundations a more complicated
than plain first-order logic), and this is the only that  nontrivial
mathematics can be formalized. But when studying relative interpretations, one
usually assumes a spartan syntax and defines relativization by
induction of formulas of the more succinct language. The approach
taken in \session{ZF-Constructible} is to consider relativizations of
the formulas (of type $\tyo$) that define each concept. For instance,
in the case of unions, we find a relativization
$\isatt{big{\uscore}union}:: (\isatt{i} \fun \isatt{o}) \fun \isatt{i}
\fun \isatt{i} \fun \isatt{o}$ of the statement
“$\union A = z$”:
\[
 \isatt{big{\uscore}union}(M,A,Z) \equiv \forall x[M].\ x \in z
 \longleftrightarrow (\exists y[M].\ y \in A \land x \in y)
\]
where $\forall x[M]\dots$ stands for $\forall x.\ M(x)\limp \dots$,
etc. The need to work with \emph{relational} presentations of defined
concepts stems from the fact that the model-theoretic definition of
$L$ requires working with set models and satisfaction, which is
defined for codes of formulas.

>>>>

The type of propositions $\prop$ is equipped with a binary operation
${\implies}$ (“meta-implication”) and a universal “meta-quantifier”
$\ALL$, that are used to represent the object
logic rules. As an example, the axiomatization of first-order logic
postulates a type $\tyo$ of booleans, and Modus Ponens
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#IFOL.mp|axiom
is written as
\begin{equation}\label{eq:modus-ponens}
  \ALL P\,Q.\ \ [P\limp Q] \implies ([P] \implies [Q]).
\end{equation}
The square brackets (which are omitted in Isabelle theories, as well
as the outermost universal quantifiers) represent
an injection from $\tyo$ into $\prop$. % ($[P]$ can be read as “$P$
% holds”)
A consequence of this representation is that every formula of
the object logic appears atomic to $\Meta$.

@@

Paulson  \cite{Paulson1989} carried out a proof that the encoding
$\Meta_{\mathrm{IFOL}}$ of
intuitionistic first-order logic IFOL without equality  in the original $\Meta$ is
conservative (there is a correspondence between provable $\phi$ in
IFOL and provable $[\phi]$ in $\Meta_{\mathrm{IFOL}}$) by putting
$\Meta_{\mathrm{IFOL}}$ proofs in \emph{expanded normal form}
\cite{MR0387024}; atomicity as stated after
Equation~(\ref{eq:modus-ponens}) plays a role in this argument. Passing to classical logic does not present
difficulties, but the addition of meta-equality must be taken care of.
Even more so, since the treatment of equality differs between the
original and the present incarnation of $\Meta$; details for the
latter are exhaustively expounded in the recent formalization by
Nipkow and Roßkof \cite{10.1007/978-3-030-79876-5_6}.

@@

For the most part, the development of set theory in Isabelle is
carried out using its ZF object logic
\cite{DBLP:journals/jar/PaulsonG96}, which is the third logical layer
of the formalization and the most versatile one, since 
Isabelle's native automation is available at this level. Apart from
the type and sort
declarations detailed above, it features a finite axiomatization,
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.mem|const
with a predicate for membership, constants for the empty set and an
infinite set, and functions $\isatt{Pow}::\tyi\fun\tyi$,
$\union::\tyi\fun\tyi$, and $\isatt{PrimReplace} :: \tyi \fun (\tyi
\fun \tyi \fun\tyo)\fun \tyi$ (for Replacement). The Axiom of
Replacement
% https://isabelle.in.tum.de/dist/library/FOL/FOL/IFOL.html#ZF_Base.replacement|axiom
has a free predicate variable $P$: % $P ::\tyi\fun\tyi\fun\tyo$:
\begin{multline*}
  (\forall x \in A .\ \forall y\, z.\ P(x, y) \wedge P(x, z)
  \longrightarrow y=z) \implies \\
  b \in \isatt{PrimReplace}(A, P)
  \longleftrightarrow(\exists x \in A .\ P(x, b)) 
\end{multline*}
The restrictions on sorts described above ensure that it is not
possible that higher-order quantification gets used in $P$. The
statement of $\AC$, also uses a free higher-order variable to denote
an indexed family of nonempty sets. % but we only
% use $\AC$ for demonstration purposes.

@@

The development of relativization and absoluteness for classes $C::
\isatt{i} \fun \isatt{o}$ follows the same pattern. Each particular
concept was manually written in a relational form and relativized.
Here, the contrast between the usual way one regards $\ZF$ as
first-order theory in the language $\{\in \}$ and the mathematical
practice of freely using defined concepts comes to the
forefront. Assistants have refined mechanisms to cope with defined
concepts and the introduction of new notation (which also make their foundations more complicated
than plain first-order logic), and this is the only way that nontrivial
mathematics can be formalized. But when studying relative interpretations, one
usually assumes a spartan syntax and defines relativization by
induction of formulas of the more succinct language. The approach
taken in \session{ZF-Constructible} is to consider relativizations of
the formulas that define each concept. For instance,
in the case of unions, we find a relativization
%https://isabelle.in.tum.de/dist/library/ZF/ZF-Constructible/Relative.html#Relative.big_union_def_raw|axiom
$\isatt{big{\uscore}union}:: (\isatt{i} \fun \isatt{o}) \fun \isatt{i}
\fun \isatt{i} \fun \isatt{o}$ of the statement
“$\union A = z$”:
\[
 \isatt{big{\uscore}union}(M,A,z) \equiv \forall x[M].\ x \in z
 \longleftrightarrow (\exists y[M].\ y \in A \land x \in y)
\]
where $\forall x[M]\dots$ stands for $\forall x.\ M(x)\limp \dots$,
etc. The need to work with \emph{relational} presentations of defined
concepts stems from the fact that the model-theoretic definition of
$L$ requires working with set models and satisfaction, which is
defined for codes of formulas in the language $\{\in\}$.
